Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-09 fscd19/fscd19.tex
[ fscd19 ] normal forms of polarized lambda

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{lcl@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
  & \mbox{case on absurd neutral} \\
\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
  \to \C\,\A\,(\cext \Gamma A)
  \to \C\,\A\,(\cext \Gamma B)
  \to \C\,\A\,\Gamma
  & \mbox{case on neutral} \\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
  & \mbox{run the monad ($\tNf$ only)} \\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^\Cov$, and
$\tcase^\Cov$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A+B$ and its two branches by the context extensions $A$
and $B$, resp.  Leaves are either ($\tabort$)
labeled by a neutral term of empty type $\tyzero$,
or ($\treturn$) by an element of $\A$.  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
flattens a 2-level case tree into a single one.)
Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^\Cov$
and $\tabort^\Cov$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

>>>>

We are looking for a cover monad $\C$ that
offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{lcl@{\quad}l}
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
  & \mbox{case on absurd neutral} \\
\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
  \to \C\,\A\,(\cext \Gamma A)
  \to \C\,\A\,(\cext \Gamma B)
  \to \C\,\A\,\Gamma
  & \mbox{case on neutral} \\
\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
  & \mbox{run the monad ($\tNf$ only)} \\
\end{array}
\]
To make things concrete,
we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
of sum type $A+B$ and its two branches by the context extensions $A$
and $B$, resp.  Leaves are either ($\tabort$)
labeled by a neutral term of empty type $\tyzero$,
or ($\treturn$) by an element of $\A$.  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
flattens a 2-level case tree into a single one.)
Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

@@

Polarized lambda calculus \cite{espiritoSanto:entcs17}
divides types deeply into positive and negative.

\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    \mid \tyzero \mid P_1 + P_2
    \mid \tyone \mid P_1 \times P_2
    \mid \Thunk N
  & \mbox{positive type}
\\
  \NTy & \ni & N,M & ::= & \natom
    \mid \top \mid N_1 \ntimes N_2
    \mid P \To N
    \mid \Comp P
  & \mbox{negative type}
\\
  \HTy & \ni & X,Y & ::= & \patom \mid P
  & \mbox{hypothesis type}
\\
\end{array}
\]
Espirito Santo's version actually does not contain the positive
product $\times$, however, it appears in Zeilberger
\cite{zeilberger:PhD}.

Polarized lambda-calculus is a focused calculus, it eagerly employs
invertible rules: the introduction rules for negative types and the
elimination rules for positive types.  As a consequence of the latter,
variables are either of atomic or negative type ($\HTy$).
Contexts $\Gamma,\Delta$ are lists of $\HTy$s.
%\fbox{$X ::= \patom \mid P$}.

To add a variable of positive type $P$ to the context, we need to
break it apart until only atoms and negative bits remain.  This is
performed by maximal pattern matching, called the left-invertible
phase of focalization.
We express maximal pattern matching on $P$ as a strong functor
$\Add P$ in the category of presheaves
mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
inductively constructed by the following rules:
\begin{gather*}
  \nru{\phyp}
      {\J\,(\cext \Gamma \patom)}
      {\AddHyp {\patom} \J \Gamma}
\qquad
  \nru{\nbranch}
      {}
      {\AddHyp {\tyzero} \J \Gamma}
\qquad
  \nru{\bbranch}
      {\AddHyp {P_1} \J \Gamma \qquad
       \AddHyp {P_2} \J \Gamma }
      {\AddHyp {P_1 + P_2} \J \Gamma}
\\[2ex]
  \nru{\nhyp}
      {\J\,(\cext \Gamma N)}
      {\AddHyp {\Thunk N} \J \Gamma}
\qquad
  \nru{\nsplit}
      {\J\,\Gamma}
      {\AddHyp {\tyone} \J \Gamma}
\qquad
  \nru{\bsplit}
      {\AddHyp {P_1} {(\Add{P_2}\,\J)} \Gamma}
      {\AddHyp {P_1 \times P_2} \J \Gamma}
\end{gather*}
Note the recursive occurrence of $\Add{P_2}$ as argument to
$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
\cite{bird:nested}.  Agda supports such nested inductive types; but
note that $\Add{P}$ is uncontroversial, since it could also be defined
by recursion on $P$.   It is tempting to name $\bsplit$
``$\tjoin$'' and $\nsplit$ ``$\treturn$''
since $\Add P$ is a graded monad on the monoid $(1,\times)$
of product types;
however, this coincidence shall not matter for our further considerations.

Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
of positive type $P$ are constructed by the
following rules:
\begin{gather*}
  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
\qquad
  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk\,N}\Gamma}
\\[2ex]
  \nru{\punit}{}{\PValT{\tyone}\Gamma}
\qquad
  \nru{\ppair}
      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
      {\PValTP{P_1 \times P_2}\Gamma}
\qquad
  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
\end{gather*}
Negative hypotheses can be eliminated by projections and application
to values, as described by the following inductive family
\fbox{$\NeT N \Gamma$}:
\begin{gather*}
  \nru{\nvar}{N \in \Gamma}{\NeT N \Gamma}
\qquad
  \nru{\tprj_i}
      {\NeTP {N_1 \ntimes N_2} \Gamma}
      {\NeT {N_i} \Gamma}
\qquad
  \nru{\tapp}
      {\NeTP {P \To N} \Gamma \qquad \PValT P \Gamma}
      {\NeT N \Gamma}
\end{gather*}
% Negative hypotheses can be eliminated by projections and application
% to values $v : \V\,P\,\Gamma$ which we leave abstract in the following
% inductive definition of \fbox{$\Ne[\V] N \Gamma$}:
% \begin{gather*}
%   \nru{\nvar}{N \in \Gamma}{\Ne[\V] N \Gamma}
% \qquad
%   \nru{\tprj_i}
%       {\NeP[\V] {N_1 \ntimes N_2} \Gamma}
%       {\Ne[\V] {N_i} \Gamma}
% \qquad
%   \nru{\tapp}
%       {\NeP[\V] {P \To N} \Gamma \qquad \V\,P\,\Gamma}
%       {\Ne[\V] N \Gamma}
% \end{gather*}

% decomposing
% decomposes
% \emph{values}, meaning inhabitants of positive types.
An elimination may either result in a negative atom $\natom$ or a
positive type $P$, which can be further taken apart by pattern
matching.  Pattern matching on neutrals is organized in the
inductively defined cover monad \fbox{$\Cov[\T]$}.
\begin{gather*}
  \nru{\treturn}{\J\,\Gamma}{\CoverT \J \Gamma}
\qquad
  \nru{\tmatch}
      {\NeTP {\Comp\,P} \Gamma \qquad \AddHypP P {\Cov[\T]\J} \Gamma}
      {\CoverT \J \Gamma}
\end{gather*}

Finally, normal forms of negative types are defined as inductive
family \fbox{$\Nf N \Gamma$}.  It relies on the previous definitions
with instance $\T = \tNf$.
\begin{gather*}
  \nru{\tne}
      {\Cov[\tNf](\tNe[\tNf]\,\natom)\,\Gamma}
      {\Nf {\natom} \Gamma}
\qquad
  \nru{\tval}
      {\Cov[\tNf](\tVal[\tNf]\,P)\,\Gamma}
      {\NfP {\Comp\,P} \Gamma}
\\[2ex]
  \nru{\npair}
      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
      {\NfP {N_1 \ntimes N_2} \Gamma}
\qquad
  \nru{\tabs}
      {\AddHypP P {\tNf\, N} \Gamma}
      {\NfP {P \To N} \Gamma}
\end{gather*}
